The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-02-11 pkg/partykit/inst/guideliketest/draft/draft.Rnw
added section on computation details; further text revisions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2461 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 581da9ae784df1de924164984aff000212702893

%% packages
\usepackage{amstext,amsfonts,amsmath,amssymb,bm,thumbpdf,lmodern,hyperref}
\usepackage[all]{hypcap}
\usepackage{enumitem}

@@

\section{Introduction}\label{sec:introduction}
In many situations fitting one global model to a data set can be very challenging, 
especially if this data contains lots of different features with a strong variation 
or complex interactions. Therefore, separating the data into more homogeneous subgroups 
based on a set of covariates first can simplify the task and fitting a local model to 
each of the resulting subgroups often leads to better results. This separation of the 
data can be done by applying a tree algorithm. Many different algorithms have been developed
in this research field. While they all follow the general idea of splitting the data such 
that some objective function is optimized, they differ in their 
specific approaches to selecting a split variable and the corresponding split point. 
Some of the first tree algorithms (e.g., AID, \citealp{Morgan+Sonquist:1963}; 
CART, \citealp{Breiman+Friedman+Stone:1984}) rely on exhaustive search procedures to find both, 
the best split point and split variable in one step by directly comparing all possible split
points in all possible split variables. However, it has been shown that this is not only
computationally expensive but also biased towards split variables with many possible split
points \citep{Doyle:1973, Kim+Loh:2001}. 
Therefore, selecting a split variable in a first step and then searching for the 
best split point only within this variable in a separated second step is a more 
promising strategy as applied for example by the algorithms QUEST \citep{Loh+Shih:1997},
CHAID \citep{Kass:1980}, GUIDE \citep{Loh:2002}, CTree \citep{Hothorn+Hornik+Zeileis:2006} 
and MOB \citep{Zeileis+Hothorn+Hornik:2008}.
For the first step of selecting a split variable they all share the same basic concept of 
choosing the covariate which shows the highest association to the response variable 
based on a statistical test. While QUEST, CHAID and GUIDE employ statistical significance
tests for contingency tables, CTree applies permutation tests in a conditional inference
framework and MOB uses fluctuation tests based on asymptotic theory. All these approaches
have shown to work well for various situations, however, the specific advantages and
disadvantages of each of the testing strategies have not yet been investigated and compared
in detail.

Therefore, in this paper the focus will be put on that first step of tree algorithms,
i.e. the task of selecting the best split variable. In particular, the approach of the 
GUIDE algorithm is compared to the one of the CTree algorithm and the MOB algorithm. 
For this purpose, a unifying framework for testing strategies in unbiased model-based tree 
algorithms is presented such that each of the three strategies can be obtained by a specific 
combination of the available building blocks. In that way, the effects of each of these 
features can be investigated separately by varying the building blocks in the tool box 
provided by this general framework.

Depending on the setup different effects can be observed for the investigated building blocks.
However, none of the results contradicts the conclusion that in general using all available 
model scores is at least as efficient but mostly clearly advantageous over considering only 
residuals. Moreover, the effects of binarizing residuals or scores at $0$ or categorizing
the values of possible split variables are investigated.
Additionaly, the impact of applying either a pre- or postpruning strategy is analyzed.

In the first part of Section~\ref{sec:urp} a basic scheme on how to build a tree model is given. 
Then, after explaining the investigated algorithms CTree, MOB, and GUIDE more in detail in 
Sections~\ref{sec:ctree}, \ref{sec:mob} and \ref{sec:guide}, the unifying framework for 
testing strategies in model-based unbiased recursive partitioning algorithms is presented in 
Section~\ref{sec:unify}.
The setting for the simulation study is introduced in Section~\ref{sec:simulation}
and the results are discussed in Section~\ref{sec:results}.

@@

\subsection{Generic Algorithm}
The basic idea of building a regression tree model is to partition the data into smaller and 
more homogeneous subgroups based on a set of covariates. Various different tree algorithms 
have been developed, all following the same general structure starting at the root of the tree
which is simply the first node containing all available data:
\begin{enumerate}
\item For the whole underlying data set in the current node some form of discrepancy 
measure is calculated. 
\item Among all available covariates one is selected as split variable and a value 
within the range/measurable space of this variable is chosen as split point such that 
after splitting the data into subgroups the discrepancy measure is minimal within the 
resulting nodes.
\item Steps 1 and 2 are repeated within each new node until either a stopping criterion
is reached or the data in the node consists of only one or all equal observations. 
\end{enumerate}

The employed discrepancy measure can for example be the sum of deviations from a typical 
or average value, but it can also be constructed based on a parametric model. For instance,
different types of residuals or their signs can be applied \citep[see e.g.,][]{Loh:2002} as 
well as rank sums or logrank scores as in \cite{Hothorn+Hornik+Wiel:2006}.
%(In the simulation study of this paper trees with linear models fitted in each node are 
%investigated with the residual sum of squares (RSS) employed as discrepancy measure.)

As explained in Section~\ref{sec:introduction} the considered tree algorithms CTree, MOB, and 
GUIDE first select the split variable and then the split point in a separate step. 
For the first step of selecting a split variable they all apply statistical tests
following the same general strategy: 
\begin{enumerate}
\item Based on the discrepancy measure an $n \times k$-matrix $D$ is obtained giving information
on the goodness of fit of the model for a data set of $n \in \mathbb{N}$ observations in the 
current node and $k \in \mathbb{N}$.
\item Using this matrix, the algorithm-specific test statistic is evaluated for each possible 
split variable $Z_j$, $j \in \{1,\ldots,m\}$ for $m \in \mathbb{N}$, assessing the dependency
of $Z_j$ on $D$.
\item For the resulting values of the test statistics the corresponding $p$-values are calculated
in order to allow for their comparison on a standardized/unified scale and in that way for
an unbiased variable selection.
%As the variables are usually not all measured at the same scale the resulting
%values of the test statistics can not be compared directly in an unbiased way. Therefore, it is
%necessary to calculate the corresponding $p$-values in order to allow for an unbiased
%variable selection on this standardized/unified scale. 
\item The variable corresponding to the smallest $p$-value is selected as split variable as 
it has shown to be the variable with the strongest influence on the goodness of fit of the model.
\end{enumerate}

@@

The CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} is based on the idea of providing 
non-parametric regression tree models in a conditional inference framework by applying 
permutation tests. 
To select a split variable it is tested whether there is any association between the transformed 
response $h(Y)$ and $g(Z)$ for a so-called influence function $g$ and each possible split 
variable~$Z$.
The function $h$ has to depend on $Y$ in a permutation-symmetric way but can be of various forms
and may also be multidimensional. For a numeric response the identity function $h(Y) = Y$ is a common 
choice, while a categorical response can be mapped to a unity vector by an indicator function 
$h(Y) = (0,\ldots,1,\ldots,0)^{\top}$.
Alternatively, the function $h$ can return a location and scale parameter of $Y$, 
e.g. $h(Y) = (Y, (Y-\bar{Y})^2)^{\top}$. 
If a parametric model is fit to the values of the response variable~$Y$ and regressor 
variable(s)~$X$ the function $h$ can be a model-based transformation returning the 
corresponding model scores by setting $h(Y) = s(Y,X,\hat{\beta})$
where $\hat{\beta}$ is the estimated model parameter and $s$ is the score function
$$
s(Y, X, \hat{\beta}) = \frac{\partial \psi(Y, X, \hat{\beta})}{\partial \beta}
$$
with the sum of $\psi$ over all observations being the objective function that is optimized 
to estimate the model parameter $\beta$. For this purpose many different M-type estimators can 
be applied including maximum likelihood or ordinary least squares. Therefore, these objective 
functions are also possible options to obtain a model-based discrepancy measure for the tree 
algorithm.

Applying the score function $s$ on a data set of $n$ observations containing the 
response vector $\boldsymbol{y} = (y_1,\ldots,y_n)^\top$ and 
regressor(s) $\boldsymbol{x} = (x_1,\ldots,x_n)^\top$ and 
on a $k$-dimensional estimated parameter vector 
$\hat{\boldsymbol{\beta}} = (\hat{\beta_1},\ldots,\hat{\beta_k})^\top$ 
yields a $n \times k$-matrix $S$.
In that way a score value indicating how well the model fits the data is given separately
for each tuple of observation and estimated parameter element 
$(y_i, x_i, \hat{\beta_l})$ with $i \in \{1,\dots,n\}$
and $l \in \{1,\ldots,k\}$.

@@

The testing strategy for selecting a split variable is based on a fitted model, i.e. on 
an estimated parameter (vector) $\hat{\beta}$ for each node of the tree, and on the 
application of the central limit theorem. 
The fitted model can be of a very basic and simple form such as an average value of the response 
vector but can also be a more complex parametric model.
In \cite{Zeileis+Hothorn+Hornik:2008} generalized linear models (GLMs) and survival regression models 
are used to exemplify the algorithm. Subsequently various other types of models have been applied as well,
for example beta regression models \citep{Gruen+Kosmidis+Zeileis:2012}, 
psychometric models such as Rasch trees \citep{Strobl+Kopf+Zeileis:2015} and
mixed effects models \citep{Fokkema+Smits+Zeileis:2018}.

Regardless of the applied model type the algorithm only requires the corresponding model 
score matrix $S$ in order to test for any instability in the model parameters over each 
particular ordering induced by each of the possible partitioning variables $Z$.
If a variable $Z$ has no specific influence on the parameters the scores would always 
fluctuate randomly around their mean 0. Therefore, these deviations are investigated by 
measuring them using an empirical fluctuation process. Applying a scalar function which 
captures the fluctuation in this process yields the test statistic of this so called 
generalized M-fluctuation test \citep{Zeileis+Hornik:2007}. 
To obtain the distribution of the resulting test statistic asymptotic theory, in particular 
the central limit theorem is applied. In this parametric unconditional framework $p$-values 
can be calculated and the covariate with the smallest $p$-value is selected as split variable.

@@

\subsection{GUIDE}\label{sec:guide}
Similar to the MOB algorithm, parametric models are fit to the nodes of a tree in 
the GUIDE algorithm \citep{Loh:2002}. Again, simple models such as 
a constant fit where the parameter $\beta$ is the mean of the response, hence 
$\hat{\beta} = \bar{y} = \frac{1}{n}\sum_{i=1,\ldots,n}^n y_i$ for a response vector 
$\boldsymbol{y} = (y_1,\ldots,y_n)^\top$, can be employed as well as more complex models.
For the statistical test applied to select a split variable some type of model-based
residuals is required. In \cite{Loh:2002} linear models are used as an example where the
residuals are defined by
$$
r(Y,X,\hat{\beta}) = Y - \hat{\beta_0} - \hat{\beta_1} \cdot X.
$$
In related literature other models together with an appropriate choice of residuals
have been applied such as in
regression trees for longitudinal and multiresponse data \citep{Loh+Zheng:2013}, 
%(http://www.stat.rice.edu/~jrojo/4th-Lehmann/slides/Loh.pdf)
quantile regression models \citep{Chaudhuri+Loh:2002} 
and proportional hazards modeling via Poisson regression \citep{Loh+He+Man:2015} among many others.
%Poisson regression trees \citep{Chaudhuri+Lo+Loh:1995}, %(adjusted Anscombe residual (Pierce and Schafer 1986))

@@

Even though CTree, MOB, and GUIDE differ in the specific tests they apply (see second column of
Table~\ref{tab:combinations}) their approaches on how to select a split variable follow the same
basic structure. Thus they can be presented together in a unifying framework for testing strategies
in model-based unbiased recursive partitioning. 
In this framework each specific testing strategy can be obtained by combining the provided 
building blocks accordingly. 
This general toolbox covers all features of the here discussed testing strategies CTree, MOB, and
GUIDE. Furthermore, it allows for all possible combinations of these features and in that way also
for variations of the commonly known testing strategies. Moreover, by exchanging only one building
block its impact can be investigated in detail.
These building blocks are explained in Section~\ref{sec:buildingblocks} and their setting
for CTree, MOB, and GUIDE are listed in Table~\ref{tab:buildingblocks} which gives a more 
detailed overview of the differences in the second column of Table~\ref{tab:combinations}.

\subsection{Building Blocks of Testing Strategies}\label{sec:buildingblocks}
The three following building blocks all refer to the way how the given data is prepared/transformed
before employing a statistical test to select a split variable.
\begin{itemize}
\item  Transformations of Y:\\
A model-based transformation of $Y$ is applied to obtain a discrepancy measure giving 
information about the goodness of fit of the model. This can be residuals $r(Y,X,\hat{\beta})$
as in the GUIDE algorithm or scores $s(Y,X,\hat{\beta})$ as in the MOB algorithm or, 
more generally, a function $h(Y,X)$ as in the CTree algorithm including both of the foregoing 
cases.
By including all available scores rather than just residuals the goodness of fit of the model is 
investigated regarding each parameter instead of just the mean/location parameter. 
Therefore, this facilitates the detection of changes in parameters other than the mean/location
parameter.

@@

\subsection{Linear Model Tree}\label{sec:lmtrees}
As an example a linear model tree is considered as this allows 
for an easy and well structured description of the investigated subjects/issues. 
For that reason, this model is also considered for the simulation study in 
Section~\ref{sec:simulation}.
However, the conclusions drawn from this example also apply for a wide range of other model
types. The models fitted in the nodes of the considered linear model tree are of the form
$$
Y = \beta_0 + \beta_1 \cdot X + \epsilon
$$
for a response variable $Y$, a regressor variable $X$ and an error $\epsilon$. 
As for each node the type of the fitted model is the same no node-specific features are considered
and the observations of the current (arbitrarily chosen) node are denoted by 
$\{(y_i,x_i)\}_{i=1,\ldots,n}$ for $n \in \mathbb{N}$.

A very basic form of a discrepancy measure can be obtained for each observation by simply 
calculating deviations of the response values from the mean value within the node, for example 
by $(\bar{y} - y_i)$.
However, this includes only the response variable $Y$ but not the fitted model.
For a model-based approach residuals can be considered defined by
$$
r(y_i,x_i,\beta_0,\beta_1) = y_i - \beta_0 - \beta_1 \cdot x_i.
$$
The appropriate choice of the function $\psi$ is then
$$
\psi(y_i,x_i,\beta_0,\beta_1) = r(y_i,x_i,\beta_0,\beta_1)^2.
$$
such that $\sum_{i=1}^n \psi(y_i,x_i,\beta_0,\beta_1)$ is the objective function
which is minimized for the estimation of the coefficient vector
$\boldsymbol{\beta} = (\beta_0, \beta_1)^\top$, 
i.e. to obtain $\boldsymbol{\hat{\beta}} = (\hat{\beta_0}, \hat{\beta_1})^\top$.
The corresponding score function is defined by
$$
s(y_i, x_i, \hat{\beta_0},\hat{\beta_1}) = 
\frac{\partial \psi}{\partial \boldsymbol{\beta}}(y_i, x_i, \hat{\beta_0},\hat{\beta_1}) = 
2 \cdot r(y_i,x_i,\hat{\beta_0},\hat{\beta_1}) \cdot (1,x_i)^\top
$$
where the factor $2$ is often omitted such that the first column of the score matrix $S$ 
corresponds to the residuals. $S$ is in this case a $n \times 2$-matrix
as $\boldsymbol{\beta}$ is two dimensional. In that way a score value indicating how well the model
fits the data is given separately for each tuple of observation and coefficient $(y_i,x_i,\beta_j)$
with $i \in \{1,\dots,n\}$ and $j \in \{1,2\}$.

In this example considering only residuals, i.e. only the first column of the score matrix $S$,
corresponds to including the derivatives of $\psi$ only regarding the intercept 
parameter~$\beta_0$.
On the contrary, using full model scores means including the whole $n \times 2$ matrix $S$, hence,
the derivatives of $\psi$ regarding the intercept $\beta_0$ (first column) and the slope parameter 
$\beta_1$ (second column).

@@

\subsection{Data Generating Process}
Each generated data set consists of a response variable, one regressor, one split 
variable and nine noise variables. In particular, the variables listed in 
Table~\ref{tab:variables} together with the corresponding distributions are 
included in the data generating process.

@@

By applying the unifying framework presented in Section~\ref{sec:unify} the effect of
each building block can be investigated on its own. In Table~\ref{tab:buildingblocks} the setting
of these building blocks is listed for CTree, MOB and GUIDE in their original form.
However, in order to be able to focus on particular features rather than the whole testing
strategies other combinations as presented in Table~\ref{tab:buildingblocks}, hence variations of
the original testing strategies, are evaluated as well.

Moreover, as prepruning is employed all of the considered statistical tests are applied as significance tests answering two questions at once: 
(1)~Should a split be performed at all? 
(2)~If so, in which variable?
For the first question the $p$-values regarding all available covariates are compared 
to a predefined level of significance $\alpha = 0.05$. Only if the smallest $p$-value is smaller
than $\alpha$ a split is performed and the covariate corresponding to this $p$-value is 
selected.

\newpage
\section{Results}
\label{sec:results}

@@

The results presented in Figure~\ref{fig:3way} clearly show that binarizing scores/residuals
reduces the power of the testing strategy as it leads to higher $p$-values corresponding
to the true split variable $Z_1$. 
This effect can be observed for a central split point (left panel, $xi=0$) as well as for a
a split point closer to upper end of the considered interval (right panel, $xi=0.8$).
However, the effect of the building block \textit{Categorization} varies over these two settings.
For a central split point all three levels perform similarly. The linear selection of split variables
is slightly ahead even though the true parameter function is a step function. 
On the contrary, clear differences can be seen for the outward split point. Here, the maximal selection
performs best while the linear selection shows its disadvantage in detecting abrupt changes
and the categorized split variables lead to even higher $p$-values and thus lower power.
Overall, it depends on the situation whether a linear or a maximal selection of a split variable
is advantageous. However, categorizing possible split variables does not lead to any improvements
but even declines the performance of a testing strategy unless the true split point is close
to a break point used for the binarization.

>>>>

%% packages
\usepackage{amstext,amsfonts,amsmath,amssymb,bm,thumbpdf,lmodern,hyperref}
\usepackage[all]{hypcap}
\usepackage{enumitem}
\usepackage{color,soul}

@@

\section{Introduction}\label{sec:introduction}
In many situations fitting one global model to a given data set can be very challenging, 
especially if the data contains lots of different features with a strong variation 
or complex interactions. Therefore, separating the data into more homogeneous subgroups 
based on a set of covariates first can simplify the task and fitting a local model to 
each of the resulting subgroups often leads to better results. This separation can be 
done by applying a tree algorithm. While all of the many developed algorithms 
follow the general idea of splitting the data such that some objective function is optimized, 
they differ in their specific approaches to selecting a split variable and the corresponding 
split point. 
Some of the first tree algorithms (e.g., AID, \citealp{Morgan+Sonquist:1963}; 
CART, \citealp{Breiman+Friedman+Stone:1984}) rely on exhaustive search procedures to find both, 
the best split point and split variable in one step by directly comparing all possible split
points in all possible split variables. However, it has been shown that this is not only
computationally expensive but also biased towards split variables with many possible split
points \citep{Doyle:1973, Kim+Loh:2001}. 
Therefore, selecting a split variable in a first step and then searching for the 
best split point only within this variable in a separated second step is a more 
promising strategy as applied for example by the algorithms QUEST \citep{Loh+Shih:1997},
CHAID \citep{Kass:1980}, GUIDE \citep{Loh:2002}, CTree \citep{Hothorn+Hornik+Zeileis:2006} 
and MOB \citep{Zeileis+Hothorn+Hornik:2008}.
For the first step of selecting a split variable they all share the same basic concept of 
choosing the covariate which shows the highest association to the response variable 
based on $p$-values provided by a statistical test. While QUEST, CHAID and GUIDE employ 
statistical significance tests for contingency tables, CTree applies permutation tests in 
a conditional inference framework and MOB uses fluctuation tests based on asymptotic theory. 
All these approaches have shown to work well for various situations, however, the specific 
advantages and disadvantages of each of the testing strategies have not yet been 
investigated and compared in detail.

Therefore, in this paper the focus will be put on that first step of tree algorithms,
i.e. the task of selecting the best split variable. In particular, the approach of the 
GUIDE algorithm is compared to the one of the CTree algorithm and the MOB algorithm
by investigating the building blocks of their testing strategies in which they differ:
(1)~Transformations of the response: using residuals or full model scores
(2)~Binarization of residuals/scores
(3)~Categorization of covariates.
For this purpose, a unifying framework for testing strategies in unbiased model-based tree 
algorithms is presented such that each of the three strategies GUIDE, CTree, and MOB can be 
obtained by a specific combination of the available building blocks. In that way, the 
effects of each of these features can be investigated separately by varying the building 
blocks in the tool box provided by this general framework. 
Additionally, the impact of applying either a pre- or postpruning strategy is analyzed.

Depending on the setup different effects can be observed for the investigated building blocks.
However, none of the results contradicts the conclusion that in general using all available 
model scores is at least as efficient but mostly clearly advantageous over considering only 
residuals. Also regarding the effects of binarizing residuals or scores at $0$ or categorizing
the values of possible split variables clear recommendations can be made based on the presented 
results.

In the first part of Section~\ref{sec:urp} a basic scheme on how to build a tree model is given. 
After explaining the investigated algorithms CTree (Section~\ref{sec:ctree}), 
MOB (\ref{sec:mob}), and GUIDE (\ref{sec:guide}) more in detail the applied pruning strategies 
(\ref{sec:pruning}) are discussed.
In Section~\ref{sec:unify} the unifying framework for testing strategies in model-based unbiased 
recursive partitioning algorithms is presented.
The setting for the simulation study is introduced in Section~\ref{sec:simulation}
and the results are illustrated and discussed in Section~\ref{sec:results}.

@@

\subsection{Generic Algorithm}
\label{sec:generic_alg}
The basic idea of building a regression tree model is to partition the data into smaller 
and more homogeneous subgroups based on a set of covariates. Various different tree 
algorithms have been developed, all following the same general structure starting at 
the root of the tree which is simply the first node containing all available data:
\begin{enumerate}
\item For the whole underlying data set in the current node some form of discrepancy 
measure is calculated. 
\item Among all available covariates one is selected as split variable and a value 
within the range/measurable space of this variable is chosen as split point such that 
after splitting the data into subgroups the discrepancy measure is minimal within the 
resulting nodes.
\item Steps 1 and 2 are repeated within each new node until either a stopping criterion
is reached or the data in the node consists of only one or all equal observations. 
\end{enumerate}

The employed discrepancy measure can for example be the sum of deviations from a typical 
or average value, but it can also be constructed based on a parametric model fitted in each 
node. For instance, different types of residuals or their signs can be applied 
\citep[see e.g.,][]{Loh:2002} as well as rank sums or logrank scores as 
in \cite{Hothorn+Hornik+Wiel:2006}.
%(In the simulation study of this paper trees with linear models fitted in each node are 
%investigated with the residual sum of squares (RSS) employed as discrepancy measure.)

As explained in Section~\ref{sec:introduction} the considered tree algorithms CTree, MOB, and 
GUIDE first select the split variable and then, in a separate step, the split point. 
For the first step of selecting a split variable they all apply statistical tests
following the same general strategy: 
\begin{enumerate}
\item Based on the discrepancy measure an $n \times k$-matrix $D$ is obtained giving information
on the goodness of fit of the model for a data set of $n \in \mathbb{N}$ observations in the 
current node and the parameter $k \in \mathbb{N}$ defined by the algorithm.
\item Using this matrix the algorithm-specific test statistic is evaluated for each possible 
split variable $Z_j$, $j \in \{1,\ldots,m\}$ for $m \in \mathbb{N}$, assessing the dependency
of $D$ on $Z_j$.
\item For the resulting values of the test statistics the corresponding $p$-values are calculated
in order to allow for their comparison on a standardized/unified scale and in that way for
an unbiased variable selection.
%As the variables are usually not all measured at the same scale the resulting
%values of the test statistics can not be compared directly in an unbiased way. Therefore, it is
%necessary to calculate the corresponding $p$-values in order to allow for an unbiased
%variable selection on this standardized/unified scale. 
\item The variable corresponding to the smallest $p$-value is selected as split variable as 
it has shown to be the variable with the strongest influence on the goodness of fit of the model.
\end{enumerate}

@@

The CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} is based on the idea of providing 
non-parametric regression tree models in a conditional inference framework by applying 
permutation tests. 
To select a split variable it is tested whether there is any association between the transformed 
response $h(Y)$ and $g(Z)$ for a so-called influence function $g$ and each possible split 
variable~$Z$.
The function $h$ has to depend on $Y$ in a permutation-symmetric way but can be of various forms
and may also be multidimensional. For a numeric response the identity function $h(Y) = Y$ is a common 
choice while a categorical response can be mapped to a unity vector by an indicator function 
$h(Y) = (0,\ldots,1,\ldots,0)^{\top}$.
Alternatively, the function $h$ can return a location and scale parameter of $Y$, 
e.g. $h(Y) = (Y, (Y-\bar{Y})^2)^{\top}$. 
If a parametric model is fit to the values of the response variable~$Y$ and regressor 
variable(s)~$X$ the function $h$ can be a model-based transformation returning the 
corresponding model scores. This is done by setting $h(Y) = s(Y,X,\hat{\beta})$
where $\beta$ is the possibly multidimensional model parameter and $s$ is the score function
$$
s(Y, X, \beta) = \frac{\partial \psi(Y, X, \beta)}{\partial \beta}
$$
with the sum of $\psi$ over all observations being the objective function that is optimized 
to obtain the estimated model parameter $\hat{\beta}$. For this purpose many different M-type 
estimators can be applied including maximum likelihood or ordinary least squares such that the 
corresponding objective function can be employed as model-based discrepancy measure for the 
tree algorithm. Evaluationg the score function $s$ on a data set of $n$ observations 
%containing the 
%response vector $\boldsymbol{y} = (y_1,\ldots,y_n)^\top$ and 
%regressor(s) $\boldsymbol{x} = (x_1,\ldots,x_n)^\top$ and on
and a $k$-dimensional estimated parameter vector 
%$\hat{\boldsymbol{\beta}} = (\hat{\beta_1},\ldots,\hat{\beta_k})^\top$ 
yields a $n \times k$-matrix.
In that way a score value indicating how well the model fits the data is given separately
for each tuple of observation and estimated parameter element 
$(y_i, x_i, \hat{\beta_l})$ with $i \in \{1,\dots,n\}$
and $l \in \{1,\ldots,k\}$.

@@

The testing strategy for selecting a split variable is based on a fitted model, i.e. on 
an estimated parameter (vector) $\hat{\beta}$ for each node of the tree, and on the 
application of the central limit theorem. 
The fitted model can be of a very basic and simple form such as an average value of the response 
vector but can also be a more complex parametric model.
In \cite{Zeileis+Hothorn+Hornik:2008} generalized linear models (GLMs) and survival regression models 
are used to exemplify the algorithm. Subsequently, various other types of models have been applied as well,
for example beta regression models \citep{Gruen+Kosmidis+Zeileis:2012}, 
psychometric models such as Rasch trees \citep{Strobl+Kopf+Zeileis:2015} and
mixed effects models \citep{Fokkema+Smits+Zeileis:2018}.

Regardless of the applied model type the algorithm only requires the corresponding model 
score matrix in order to test for any instability in the model parameters over each 
particular ordering induced by each of the possible partitioning variables $Z$.
If a variable $Z$ has no specific influence on the parameters the ordered scores would 
always fluctuate randomly around their mean 0. Therefore, these deviations are investigated 
by measuring them using an empirical fluctuation process. Applying a scalar function which 
captures the fluctuation in this process yields the test statistic of this so called 
generalized M-fluctuation test \citep{Zeileis+Hornik:2007}. 
To obtain the distribution of the resulting test statistic asymptotic theory, in particular 
the central limit theorem is applied. In this parametric unconditional framework $p$-values 
can be calculated and the covariate with the smallest $p$-value is selected as split variable.

@@

\subsection{GUIDE}\label{sec:guide}
Similar to the MOB algorithm, parametric models are fit to the nodes of a tree in 
the GUIDE algorithm \citep{Loh:2002}. Again, simple models such as 
a constant fit whith the parameter $\beta$ being the mean of the response, hence 
$\hat{\beta} = \bar{y} = \frac{1}{n}\sum_{i=1,\ldots,n}^n y_i$ 
%for a response vector $\boldsymbol{y} = (y_1,\ldots,y_n)^\top$, 
can be employed as well as more complex models.
For the statistical test applied to select a split variable some type of model-based
residual is required. In \cite{Loh:2002} linear models are used as an example where the
residuals are defined by
$$
r(Y,X,\hat{\beta}) = Y - \hat{\beta_0} - \hat{\beta_1} \cdot X.
$$
In related literature other models together with an appropriate choice of residuals
have been applied such as in
regression trees for longitudinal and multiresponse data \citep{Loh+Zheng:2013}, 
%(http://www.stat.rice.edu/~jrojo/4th-Lehmann/slides/Loh.pdf)
quantile regression models \citep{Chaudhuri+Loh:2002} 
and proportional hazards modeling via Poisson regression \citep{Loh+He+Man:2015} among many others.
%Poisson regression trees \citep{Chaudhuri+Lo+Loh:1995}, %(adjusted Anscombe residual (Pierce and Schafer 1986))

@@

Even though CTree, MOB, and GUIDE differ in the specific tests they apply (see second column of
Table~\ref{tab:combinations}) their approaches on how to select a split variable follow the same
basic structure as explained in Section~\ref{sec:generic_alg}. 
Thus they can be presented together in a unifying framework for testing strategies
in model-based unbiased recursive partitioning. 
In this framework each specific testing strategy can be obtained by combining the provided 
building blocks accordingly. But apart from that, variations of the commonly known testing 
strategies can be obtained as this general toolbox allows for all possible combinations of the 
building blocks. Moreover, by exchanging only one of them its impact can be investigated in detail.
These building blocks are explained in Section~\ref{sec:buildingblocks} and their setting
for CTree, MOB, and GUIDE are listed in Table~\ref{tab:buildingblocks} which gives a more 
detailed overview of the differences in the second column of Table~\ref{tab:combinations}.

\subsection{Building Blocks of Testing Strategies}\label{sec:buildingblocks}
The three following building blocks all refer to the way how the given data is prepared/transformed
before employing a statistical test to select a split variable.
\begin{itemize}
\item  Transformations of Y:\\
A model-based transformation of $Y$ is applied to obtain a discrepancy measure giving 
information about the goodness of fit of the model. This can be residuals $r(Y,X,\hat{\beta})$
as in the GUIDE algorithm or scores $s(Y,X,\hat{\beta})$ as in the MOB algorithm or, 
more generally, a function $h(Y,X)$ as in the CTree algorithm including both of the foregoing 
cases.
By including all available scores rather than just residuals the goodness of fit of the model is 
investigated regarding each parameter instead of just the mean. 
%Therefore, this facilitates the detection of changes in parameters other than the mean/location parameter.

@@

\subsection{Linear Model Tree}\label{sec:lmtrees}
As an example a linear model tree is considered as this allows for an easy and well structured 
description of the investigated subjects. However, the conclusions drawn from this example also 
apply for a wide range of other model types. 
The models fitted in the nodes of the linear model tree as considered for the simulation 
study in Section~\ref{sec:simulation} are of the form
$$
Y = \beta_0 + \beta_1 \cdot X + \epsilon
$$
for a response variable $Y$, a regressor variable $X$ and an error $\epsilon$. 
For each node the type of the fitted model is the same such that no node-specific features are 
considered and the observations of the current (arbitrarily chosen) node are denoted by 
$\{(y_i,x_i)\}_{i=1,\ldots,n}$ for $n \in \mathbb{N}$.

A very basic form of a discrepancy measure can be obtained for each observation by simply 
calculating deviations of the response values from the mean value within the node, for example 
by $(\bar{y} - y_i)$.
However, this includes only the response variable $Y$ but not the fitted linear model.
For a model-based approach residuals can be considered defined by
$$
r(y_i,x_i,\beta_0,\beta_1) = y_i - \beta_0 - \beta_1 \cdot x_i.
$$
The appropriate choice of the function $\psi$ is then
$$
\psi(y_i,x_i,\beta_0,\beta_1) = r(y_i,x_i,\beta_0,\beta_1)^2
$$
such that $\sum_{i=1}^n \psi(y_i,x_i,\beta_0,\beta_1)$ is the objective function
which is minimized to obtain the estimated coefficient vector
$\boldsymbol{\hat{\beta}} = (\hat{\beta_0}, \hat{\beta_1})^\top$.
The corresponding score function is defined by
$$
s(y_i, x_i, \hat{\beta_0},\hat{\beta_1}) = 
\frac{\partial \psi}{\partial \boldsymbol{\beta}}(y_i, x_i, \hat{\beta_0},\hat{\beta_1}) = 
2 \cdot r(y_i,x_i,\hat{\beta_0},\hat{\beta_1}) \cdot (1,x_i)^\top
$$
where the factor $2$ is often omitted such that the first column of the $n \times 2$-dimensional
score matrix corresponds to the residuals. 
Therefore, including only residuals (i.e. only the first column of the score matrix)
corresponds to including the derivatives of $\psi$ only regarding the intercept 
parameter~$\beta_0$.
On the contrary, by using full model scores (i.e. the whole $n \times 2$ matrix) 
the derivatives of $\psi$ regarding the intercept $\beta_0$ (first column) and the slope 
parameter $\beta_1$ (second column) are considered.

@@

\subsection{Data Generating Process}
Each generated data set consists of a response variable, one regressor, one split 
variable and nine noise variables as listed in Table~\ref{tab:variables} 
together with the corresponding distributions.

@@

As explained before, the aim of this simulation study is to investigate the effects
of each particular building block of the unifying framework presented in Section~\ref{sec:unify} 
rather than the whole testing strategies. Therefore, other combinations as presented in 
Table~\ref{tab:buildingblocks}, hence variations of GUIDE, CTREE, and MOB are evaluated as well
as their original versions. Moreover, as prepruning is employed all of the considered statistical 
tests are applied as significance tests answering two questions at once: 
(1)~Should a split be performed at all? 
(2)~If so, in which variable?
For the first question the $p$-values regarding all available covariates are compared 
to a predefined level of significance $\alpha = 0.05$. Only if the smallest $p$-value is smaller
than $\alpha$ a split is performed and the covariate corresponding to this $p$-value is selected.

%\newpage
\section{Results}
\label{sec:results}

@@

The results presented in Figure~\ref{fig:3way} clearly show that binarizing scores/residuals
reduces the power of the testing strategy as it leads to higher $p$-values corresponding
to the true split variable $Z_1$. 
This effect can be observed for a central split point (left panel, $\xi=0$) as well as for a
a split point closer to upper end of the considered interval (right panel, $\xi=0.8$).
However, the effect of the building block \textit{Categorization} varies over these two settings.
For a central split point all three levels perform similarly. The linear selection of split variables
is slightly ahead even though the true parameter function is a step function. 
On the contrary, clear differences can be seen for the outward split point. Here, the maximal selection
performs best while the linear selection shows its disadvantage in detecting abrupt changes
and the categorized split variables lead to even higher $p$-values and thus lower power.
Overall, it depends on the situation whether a linear or a maximal selection of a split variable
is advantageous. However, categorizing possible split variables does not lead to any improvements
but even declines the performance of a testing strategy unless the true split point is close
to a break point used for the binarization.
